Predicting The Success of The Female United States Artistic Gymnastics Team at The 2024 Paris Olympics Using Mean and Standard Deviation Derived Parameters

### 2024-01-10 manuscript/submission.tex
committing recent changes to the abstract, introduction, and data section based on the professor's statistical writing notes

NOT DONE

@@

  Open with a sentence to establish the importance of the subject of the paper.
  Identify a gap in the literature to set up the background of the paper.
  Highlight the novelty/contributions of the paper.
  For application papers, allude to new discoveries and their impacts.
  For method papers, outline the essence of the methodology, and evidence from theoretical and numerical studies supporting the methods.
  It must make sense when read in isolation for those who only read the abstract, and must also provide a clear and accurate summary of the manuscript for readers who read the entire manuscript (Zeiger 2000).

@@

For decades, gymnastics has been the most watched sport in the Summer Olympics and the United States 
has been known for bringing gold medal-winning gymnasts to compete. In the 2020 
Summer Olympics games in Tokyo, the United States female artistic gymnastics team took home 
the gold medal in the Women's All Around, a bronze medal in the Women's Balance 
Beam, a gold medal in the Women's Floor Exercise, a silver medal for the Women's
Team, a bronze medal in the Women's Uneven Bars, and a silver medal in the Women's Vault. 
Noting that the United States was the only country to have a team medal in all six women's artistic gymnastics 
categories, the likely hood of the United States bringing female athletes that will medal in the 2024 
Paris Olympics is very probable. However, in the 2020 Summer Olympic games in Tokyo, the United States 
male artistic gymnastics team did not medal at all. For the purposes of this data challenge, I focused my 
time on finding a statistical method that best isolates strong female gymnasts that would be prime candidates 
for the 2024 olympics as the female USA team historically earns more medals than the male team. Later, I applied 
my statistical method to the male team, which I anticipate will earn fewer medals than the female team, if any. 
Noting that this data challenge can be interpreted in a number of ways, I designed my methods to maximize the 
overall medal count while also placing greatest importance on the team all-around medal. In this paper, I will 
identify which five female and male artistic gymnasts are best suited to be sent to the Paris 2024 
Olympics on behalf of the United States in order to maximize the expected medal count.

Currently there is no literature available relevant to predicting gymnastics outcomes at the 2024 
Paris Summer Olympics or at recent olympics. However, there is literature that relates to various aspects of 
the data that will be used to form my predictions. For example, the article ``The Prediction of All-Around 
Event Final Score Based On D and E Score Factors In Women's Artistic Gymnastics'' aims to determine the 
impact of individual apparatus difficulty and execution scores on overall score. 
Expertise from this article could be useful in allowing us to understand how D and E scores, 
which are available in my data, play an effort in predicting the overall score. Additionally, the paper 
``Performance rating in men’s world elite artistic gymnastics: a status-quo study on scoring 
tendencies at Olympic Games following rule changes'' also can play a similar role in predicting overall 
score. This paper's objective is to analyze the effect on elite artistic gymnastics had by the 2006 rule change 
which aimed to reduce the dominance of difficulty while strengthening performance differentiation in 
competition. Notably, the paper found that the rule change resulted in significantly enhanced performance 
differentiation with the execution or E score being the main predictor of the final score as opposed to 
the difficulty or D score. Furthermore, in the article ``How Apparatus Difficulty Scores Affect All 
Around Results in Men's Artistic Gymnastics,'' the authors use regression, cluster, and ANOVA analysis to 
analyze how difficulty scores predict overall score differently across different apparatuses. However, in this 
article the authors found that while difficulty scores may have predictive tendancies for some apparatuses, 
the strength of the prediction varied drastically from apparatus to apparatus, with there being little to no 
predictive ability for some apparatuses. Ultimately, the available literature on predicting artistic 
gymnastic scores seemed generally limited to how we can use the E and D scores provided in the data sets 
to predict overall score, but some aspects of this literature casted doubt on whether E and D scores should be 
used as predictors at all.

@@

To predict which of the United States athletes are most likely to medal on the various apparatuses, 
I will be using the cleaned data that 
is provided in the UCSAS data challenge that includes data from major domestic and international 
gymnastic competitions from 2022 and 2023. I have opted against using the data from major domestic
and international gymnastic competions from 2017 to 2021 as time, injuries, 
and other factors can have a large impact on the success of a gymnast, so the greatest predictor
of the Olympic outcomes in 2024 will be the most recent data.
\\
For the purposes of my research design, when determining the most ideal feamle olympic candidates, 
I limited the data I utilized to only include female United States
gymnasts, presuming that the United States female gymnasts have a strong chance at earning medals in any category, 
based on the results of the 2020 Summer Olympics. When determining the most ideal American male candidates, 
I did inspect the data of non United States gymnasts in order to determine which medals the United States 
male team has the greatest chance of earning. 
Furthermore, I processed the data to remove any scores of zero based on the assumption that these may 
be indicative of missing scores and that instances of these scores are rare. I further justified my decision 
to eliminate scores of zero for the purposes of knowing the true mean scores of each athlete on each 
apparatus for each successful attempt. Additionally, for some athletes and competitions, I found that the vault
apparatus was listed as ``VT,'' ``VT1,'' and ``VT2.'' This discrepancy is due to the fact that athletes are permitted
two attempts on the vault apparatus. In an effort to interpret the data as accurately as possible and avoid having 
multiple statistics for one athlete on the same apparatus, I corrected this discrepancy by merging ``VT,'' ``VT1,'' and 
``VT2'' to all be reported as ``VT.'' Also, the data source is flawed as it alternates between reporting athlete names
in upper or lower case letters. In order to correct this, I converted all last names to lower case for consistency 
purposes. Lastly, the data source is inconsistent in the manner that it reports the first names of athletes as certain
athletes are listed using only their first name at some competitions and are listed using their first and middle names 
under the ``first name'' category for other competitions. In order to resolve this discrepancy, I manually verified that 
any athletes listed under two names were the same person and I further verified that there were no United States  
athletes with the same last name so that all statistical computations could be completed through grouping by last 
name alone.

@@

>>>>

@@

  Following the 2020 Toyko Olympics where the United States was the only country to have a team medal 
  in all six of the women's artistic gymnastic events, the performance of the United States women's artistic 
  gymnastics team at the 2024 Paris Summer Olympics is once again predicted to be the most watched Olympic 
  dicipline both in the United States and globally. With the whole world anticipating the success of the United 
  States women's team, it is crucial that the team of five women that are selected to represent the United States 
  are the athletes best suited to optimize success and maximize the expected medal count in Paris 2024. As there 
  are no preexisting statistical papers that methodize the selection of athletes for Olympic Gymnastic teams, 
  this paper aims to answer the question of what metric should be used to judge the olympic canidacy of USA 
  gymnasts and to outline a statistical method that can be used presently and in the future to select the ``best'' 
  gymnasts to comprise the United States women's Olympic team in order to maximize success. Using an anatlytical 
  method dependent on mean and standard deviation derived parameters, this paper outlines a procedure for identifying 
  the group of five athletes who will enable the Team USA Olympic Women’s Artistic Gymnastics team to optimize 
  success through numberically evaluating each USA athlete based on both their accuracy and precision in order to 
  more accurately identify the ``best'' team candidates. Through application of this analytical method to present 
  USA female artistic gymnasts, Simone Biles, Shilese Jones, Konnor McClain, Kaliya Lincoln, and Jade Carey can be 
  identified as the the group of five athletes who will enable the Team USA Olympic Women’s Artistic Gymnastics 
  teams to optimize success at the 2024 Paris Summer Olympics.

@@

At the Summer Olympics in Paris in 2024, ninety-six men and ninety-six women from around the world will 
represent their countries and compete in the artistic gymnatics dicipline. The overall importance of the Olympics 
cannot be understated as the games are a symbol of peaceful global interaction and give people hope that a better 
world is possible. For decades, gymnastics has been the most watched sport in the Summer Olympics and the United 
States has been known for bringing gold medal-winning gymnasts to compete. In the 2020 
Summer Olympics games in Tokyo, the United States women's artistic gymnastics team was the only team to take 
home a medal in all six categories. Hence, the likelihood of the USA Olympic Women’s Artistic Gymnastics 
team bringing female athletes that will medal in the 2024 Paris Olympics is very probable. 
However, in the 2020 Summer Olympic games, the United States 
male artistic gymnastics team did not medal at all. As the female USA team historically earns more medals than 
the male team, the focus of this paper is aimed at identifying a method that best isolates strong female gymnasts 
that would be prime candidates for the 2024 and future Olympics. Later in this paper, the designed statistical 
method is applied to the male team, which is anticipated to earn fewer medals than the female team, if any 
(see Appendix). Noting that Olympic ``success'' can be interpreted in a number of ways, the methods of this 
paper are designed to maximize the overall medal count by maximizing the USA team's opportrunities to medal in each 
category while also placing greatest importance on the team all-around medal.

\\

Currently there is no literature available relevant to predicting gymnastics outcomes at the 2024 
Paris Summer Olympics or at recent olympics. However, there is literature that relates to various aspects of 
the data that will be used to form my predictions. For example, the article ``The Prediction of All-Around 
Event Final Score Based On D and E Score Factors In Women's Artistic Gymnastics'' aims to determine the 
impact of individual apparatus difficulty and execution scores on overall score. 
Expertise from this article could be useful in allowing us to understand how difficulty and execution scores 
play a role in predicting the overall score. Additionally, the paper 
``Performance rating in men’s world elite artistic gymnastics: a status-quo study on scoring 
tendencies at Olympic Games following rule changes'' also can play a similar role in predicting overall 
score. Notably, the paper found that recent Olympic rule changes resulted in significantly enhanced performance 
differentiation with the execution score being the main predictor of the final score. 
Furthermore, in the article ``How Apparatus Difficulty Scores Affect All 
Around Results in Men's Artistic Gymnastics,'' the authors use regression, cluster, and ANOVA analysis to 
analyze how difficulty scores predict overall score differently across different apparatuses. However, in this 
article the authors found that while difficulty scores may have predictive tendancies for some apparatuses, 
the strength of the prediction varied drastically from apparatus to apparatus, with there being little to no 
predictive ability for some apparatuses. Ultimately, the available literature on predicting artistic 
gymnastic scores seemed generally limited to how difficulty and execution scores predict overall score, 
but some aspects of this literature casts doubt on whether difficulty and execution scores should be 
used as predictors at all.

\\

As existing literature fails to suggest how data from major domestic and international gymnastics competitions 
from the seasons leading up to the Summer Olympics can be used to identify the ``best'' Olympic team candidates, 
the analytical method presented in this paper aims to outline a procedure that takes into account data from recent 
gymnastics competitions and Olympic competition structure in orer to identify the group of five athletes who will 
enable the Team USA Olympic Women’s Artistic Gymnastics team to optimize success in Paris 2024. If Olympic 
competition structure and recent competitive scores on each apparatus are taken into consideration, then an 
anatlytical method dependent on mean and standard deviation derived parameters can be used to identify the foremost 
candidates to compete on behalf of the United States at the Olympics.

\\

The rest of this paper is organized as follows. First, I will evaluate the data that is available in order to complete 
this project in Section~\ref{sec:data}. Then, I will discuss my 
analytical methods to determine the atheltes that should be sent to the Olympics from the United States 
in Section~\ref{sec:meth}. Later, I will present the results of my 
methods in Section~\ref{sec:res}. Lastly, I will discuss the meaning of my results, some challenges, 
and some limitations of this project in Section~\ref{sec:dis}.

@@

To predict which of the United States athletes are most likely to medal on the various apparatuses, 
I will be using the cleaned data that is provided by the UConn Sports Analytics Symposium data challenge 
that includes data from major domestic and international gymnastic competitions from 2022 and 2023. 
I have opted against using the older data from 2017 to 2021 as time, injuries, 
and other factors can have a large impact on the success of a gymnast, so the greatest predictor
of the Olympic outcomes in 2024 will be the most recent data.
\\
For the purposes of my research design, when determining the most ideal feamle olympic candidates, 
I limited the data I utilized to only include female United States
gymnasts, presuming that the United States female gymnasts have a strong chance at earning medals in any category, 
based on the results of the 2020 Summer Olympics. 
Furthermore, I processed the data to remove any scores of zero based on the assumption that these may 
be indicative of missing scores and that instances of these scores are rare. I further justified my decision 
to eliminate scores of zero for the purposes of knowing the true mean scores of each athlete on each 
apparatus for each successful attempt. Additionally, for some athletes and competitions, I found that the vault
apparatus was listed as ``VT,'' ``VT1,'' and ``VT2.'' This discrepancy is due to the fact that athletes are permitted
two attempts on the vault apparatus. In an effort to interpret the data as accurately as possible and avoid having 
multiple statistics for one athlete on the same apparatus, I corrected this discrepancy by merging ``VT,'' ``VT1,'' and 
``VT2'' to all be reported as ``VT.'' Also, the data source is flawed as it alternates between reporting athlete names
in upper or lower case letters. In order to correct this, I converted all last names to lower case for consistency 
purposes. Lastly, the data source is inconsistent in the manner that it reports the first names of athletes as certain
athletes are listed using only their first name at some competitions and are listed using their first and middle names 
under the ``first name'' category for other competitions. In order to resolve this discrepancy, I manually verified that 
any athletes listed under two names were the same person and I further verified that there were no United States  
athletes with the same last name so that all statistical computations could be completed through grouping by last 
name alone.

@@

When determining the most ideal American male candidates, 
I did inspect the data of non United States gymnasts in order to determine which medals the United States 
male team has the greatest chance of earning.
